---
title: "RNA-Seq Analysis with STAR"
output: html_document
date: "2026-05-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Running the DESeq2 Analysis.

Load the necessary libraries for the analysis.
```{r}
library(tidyverse)
library(tximport)
library(DESeq2)
library(readr)
library(matrixStats)
library(ggplot2)
library(limma)
```

Set the sample names and the conditions.
### CHANGE SAMPLE NAMES AND CONDITIONS HERE FOR DIFFERENT ANALYSES ###

```{r}
# Add in the sample names as a list below
sample_names = c('input sample names here')
# Add in the conditions below with the correct number of replicates for each condition
sample_condition = c(rep('Condition 1',3),rep('Condition 2',3),rep('Condition 3',3))
unique_condition = unique(sample_condition)
num_conditions = length(unique_condition)

```


Define the abundance files paths and import genes counts text files 
### CHANGE PATHS HERE FOR DIFFERENT ANALYSES ###

```{r}
analysis_path = "/path/to/STAR/outputs"
abundance_path = file.path(analysis_path, sample_names,"gene_feature_counts.txt")
names(abundance_path) = sample_names

# Convert list of count file paths to a list of dataframes within those paths
count_list = lapply(abundance_path, function(f) {
  df = read.table(f, header=TRUE, comment.char="#")
})
# Convert the count list to only contain the gene counts from the 7th column of the count text files
for(c in 1:length(count_list)){
  names(count_list[[c]])[7] = sample_names[c]
  count_list[[c]] = count_list[[c]][,c(1,7)]
}

# merge the lists in count_list by GeneID
countdata = Reduce(function(x, y) merge(x, y, by=1), count_list)

# set rownames as the GeneID
rownames(countdata) = countdata$Geneid
countdata = countdata[,-1]

# rename columns as the sample_names
colnames(countdata) = sample_names
```

Define the sample dataframe and make a DESeq2 dataframe. 

```{r}
sample_df = data.frame(sample = factor(sample_names),
              condition = factor(sample_condition,levels = unique_condition ))
# set row names attribute of sample_df
row.names(sample_df) = sample_names

dds = DESeqDataSetFromMatrix(
  countData = countdata,
  colData = sample_df,
  design = ~ condition
)
```

filter out low count genes 

```{r}
# pre-filter low count genes
keep = rowSums(counts(dds)) >= 10
dds = dds[keep,]
```

Relevel and make multiple dds objects if there are more than 2 conditions

```{r}
# COPY AND PASTE THE RELEVELING OF DDS DONE BELOW BASED ON WHICH CONDITIONS YOU WANT TO SEE AS THE REFERENCE CONDITION 

dds$condition = relevel(dds$condition, ref = "x")
dds_x = DESeq(dds)

# COPY AND PASTE THE REFACTORING OF DDS DONE BELOW BASED ON THE ORDER OF CONDITIONS YOU WANT TO SEE IN DOWNSTREAM ANALYSES 

dds$condition = factor(dds$condition, levels = c("Condition 1", "Condition 2","Condition 3"))

```

Extract the results object and perform lfc shrinkage operations

```{r}
# COPY AND PAST THE LFC SHRINKAGE TO REPEAT THIS OPERATION BASED ON COMBINATIONS OF CONDITIONS YOU WANT COMPARED IN YOUR ANALYSIS
# Reference = x
res_x_y = results(dds_s, contrast = c('condition','y','x'),alpha = 0.05)
res_x_y.lfcShrink = lfcShrink(dds_x,
                    res = res_x_y,
                    coef = 'condition_y_vs_x',
                    type = 'apeglm')

```

Get Gene symbol names for each ENSG ID:

```{r}
library(biomaRt)
```

Rid of version numbers in ENSG IDs and use Biomart to get Gene names. 

```{r}
ensembl_names = row.names(dds_x)
# rids of version numbers after a "." character in ENSG IDs
ensembl_names_no_version = gsub("\\.\\d+$", "", ensembl_names)

# searches biomart for external gene names 
ensembl = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

gene_info = getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                       filters = "ensembl_gene_id",
                       values = ensembl_names_no_version,
                       mart = ensembl)

```

```{r}
# create a vector of gene names aligned with res
gene_names = gene_info$external_gene_name[match(ensembl_names_no_version, gene_info$ensembl_gene_id)]

# replace NAs with the Ensembl ID itself
gene_names[is.na(gene_names) | gene_names == ""] = ensembl_names[is.na(gene_names)]

```

Give every res object a column with external gene name

```{r}
res_x_y.lfcShrink$geneID = gene_names
```

Function for volcano plot and results 

```{r}
volcano_plot = function(res,lfc_val,comp_name){
  
  res$log2FoldChange = as.numeric(res$log2FoldChange)
  # Keep NA as NA, do not replace
  res$neglog_padj = -log10(res$padj)
  # Optionally, remove rows with NA if you want to completely exclude them from plotting
  res_plot = res[!is.na(res$neglog_padj), ]
  res_plot$significant = ifelse(!is.na(res_plot$padj) & res_plot$padj < 0.0005, "Significant", "Not Significant")  

  plot_labels = res_plot[!is.na(res_plot$padj) & abs(res_plot$log2FoldChange) > lfc_val & res_plot$padj < 0.0005, ]
  
  volcano_plot = ggplot(res_plot, aes(x = log2FoldChange, y = neglog_padj, color = significant)) +
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.0005), linetype = "dotted", color = "blue") +
    geom_vline(xintercept = lfc_val, linetype = "dotted", color = "blue") +
    geom_vline(xintercept = lfc_val*(-1), linetype = "dotted", color = "blue") +
    geom_text(data = plot_labels, aes(label = geneID, vjust = -0.5), size = 2) +
    annotate("text", x = Inf, y = Inf, label = paste0(length(res_plot$geneID)," genes present"), hjust = 2, vjust = 1.1, size = 3) +
  xlim(-10,10)+
  labs(title = paste0("Volcano Plot for ", comp_name), x = "Log2 Fold change", y = "-log(adjusted p-value)", 
       color = "Significance of Differential Expression")
  
  save_name = gsub(" ","_",comp_name)
    # Saves the volcano plot to a directory
  setwd(analysis_path)
  ggsave(paste0(save_name,"_volcanoplot_TODAYSDATE.pdf"), plot = volcano_plot, width = 10, height = 6, dpi = 300)
  
  res_top_genes = res[!is.na(res$padj) & abs(res$log2FoldChange) > 1 & res$padj < 0.05, ]
  res_top_genes_sorted = res_top_genes[order(res_top_genes$padj,decreasing = FALSE),]
  res_top_10 = head(res_top_genes_sorted,10)
  res_top_10 = as.data.frame(res_top_10)
  top_file_name = paste0(comp_name,collapse = "")
  write_csv(res_top_10,file = paste0(top_file_name,"_top10_genes.csv"))
  
  
  return(volcano_plot)
  
}
```

## Heat maps for specific genes

```{r}
library(pheatmap)
```

Function for generating heatmap based on genes of interest

```{r}
heatmap = function(dds, genes, gene_names, plot_name,anno_rows,logvals,displayNum = FALSE){
  
    # Set rownames of dds to gene names (important if needed for indexing)
    rownames(dds) = gene_names
  
  # Filter genes to only those present in gene_names
  valid_genes = intersect(genes, gene_names)

  if (length(valid_genes) == 0) {
    stop("None of the input genes are present in gene_names.")
  }
  
  # Get size-factor normalized counts
  normvalues = counts(dds, normalized = TRUE)

  # Subset to valid genes only
  gene_vals = normvalues[valid_genes, , drop = FALSE]
  plot_name = paste0(plot_name, " Raw Expression")
  save_name = paste0(gsub(" ","_",plot_name),"_Raw_heatmap_TODAYSDATE.pdf")
  
  if (logvals){
    gene_vals = log2(gene_vals + 1)
    plot_name = paste0(plot_name, " Log2 Expression")
    save_name = paste0(gsub(" ","_",plot_name),"_Log2_heatmap_TODAYSDATE.pdf")
  }
  
  print(nrow(gene_vals))
  
  annotation_col = as.data.frame(colData(dds)[, "condition", drop = FALSE])
  
  if (is.logical(anno_rows)){
    anno_rows_df = NULL
  } else {
    anno_rows_df = data.frame(Gene_Type = anno_rows)
    rownames(anno_rows_df) = rownames(gene_vals)
  }
  
  displayNum_val = if (displayNum) round(gene_vals,0) else FALSE

  # Plot
  setwd(analysis_path)
  pdf(save_name, width = 50, height = 30)

  pheatmap::pheatmap(
    gene_vals,
    show_rownames = TRUE,
    show_colnames = FALSE,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    display_numbers = displayNum_val,
    cellwidth = 50,
    cellheight = 20,
    fontsize = 20,
    border_color = NA,
    main = plot_name,
    annotation_col = annotation_col,
    annotation_row = anno_rows_df
  )

  dev.off()
}
```

Heat map for Pearson and Spearman correlation test

```{r}
Pearson_Spearman_heatmap = function(dds,upper_color,lower_color){
  vsd = vst(dds, blind = TRUE)

  mat = assay(vsd)
  
  Pearson_cor = cor(mat)
  Pearson_cor[upper.tri(Pearson_cor)] = NA
  
  Spearman_cor = cor(mat,method = "spearman")
  Spearman_cor[upper.tri(Spearman_cor)] = NA
  
  annotation_col = as.data.frame(colData(dds)[, "condition", drop = FALSE])
  
  setwd(analysis_path)
  pdf("Pearson_Heatmap_TODAYSDATE.pdf", width = 15, height = 15)
  
  pheatmap(Pearson_cor, 
        show_rownames = FALSE,
        show_colnames = FALSE,
        cluster_rows = FALSE,
        cluster_cols = FALSE,
        cellwidth = 20,
        cellheight = 20,
        fontsize = 20,
        na_col = "white", 
        colorRampPalette(c(lower_color, upper_color))(100),
        main = "Pearson Correlation Heatmap",
        annotation_col = annotation_col,
        annotation_row = annotation_col)

  dev.off()

  pdf("Spearman_Heatmap_TODAYSDATE.pdf", width = 15, height = 15)
  
  pheatmap(Spearman_cor, 
        show_rownames = FALSE,
        show_colnames = FALSE,
        cluster_rows = FALSE,
        cluster_cols = FALSE,
        cellwidth = 20,
        cellheight = 20,
        fontsize = 20,
        na_col = "white", 
        colorRampPalette(c(lower_color, upper_color))(100),
        main = "Spearman Correlation Heatmap",
        annotation_col = annotation_col,
        annotation_row = annotation_col)

  dev.off()
  
}

```


Perform PCA and plot. This will include the top 1000 variance genes in the analysis. 

```{r}
PCA_plotter = function(dds,cond_colors){
  vsd = vst(dds, blind = TRUE)

  mat = assay(vsd)
  
  # pick top variable genes
  vars = rowVars(mat)
  top = order(vars, decreasing = TRUE)[1:1000]
  # run PCA
  pca = prcomp(t(mat[top, ]), center = TRUE, scale. = TRUE)
  loadings = pca$rotation
  
  pca_df = as.data.frame(pca$x)
  pca_df$sample = rownames(pca_df)
  
  # add metadata
  pca_df = merge(pca_df, as.data.frame(colData(vsd)), by = "sample")
  
  percentVar = (pca$sdev^2) / sum(pca$sdev^2)
  
  PCA_Plot = ggplot(pca_df, aes(PC1, PC2, color = condition)) +
    geom_point(size = 3) +
    scale_color_manual(values = cond_colors) +
    xlab(paste0("PC1: ", round(percentVar[1] * 100), "%")) +
    ylab(paste0("PC2: ", round(percentVar[2] * 100), "%")) +
    theme_bw()

  setwd(analysis_path)
  ggsave(paste0("PCA_plot_TODAYSDATE.pdf"), plot = PCA_Plot, width = 10, height = 6, dpi = 300)
  
  print(PCA_Plot)
  
  top_PC1 = head(order(abs(loadings[, "PC1"]), decreasing = TRUE), 100)
  genes_PC1 = rownames(loadings)[top_PC1]
  write.csv(genes_PC1, "genes_PC1.csv", row.names = FALSE)
  
  top_PC2 = head(order(abs(loadings[, "PC2"]), decreasing = TRUE), 100)
  genes_PC2 = rownames(loadings)[top_PC2]
  write.csv(genes_PC2, "genes_PC2.csv", row.names = FALSE)
  
}
```

Comparative Bar Graph

```{r}
comp_bar_graph = function(res,genes,plot_name) {
  
  all_genes = res$geneID

  valid_genes = intersect(genes, all_genes)

  if (length(valid_genes) == 0) {
    stop("None of the input genes are present in gene_names.")
  }

  res_genes = res[res$geneID %in% valid_genes, ]

  plot_df = res_genes[, c("geneID", "log2FoldChange")]
  
  com_bar = ggplot(plot_df, aes(x = geneID, y = log2FoldChange)) +
  geom_bar(stat = "identity") +
  geom_point(size = 2) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) +
  ylab("Log2 Fold Change") +
  xlab("Gene") +
  ggtitle(paste0("Comparative Bar Plot for ", plot_name))
  
  save_name = gsub(" ","_",plot_name)
  setwd(analysis_path)
  ggsave(filename = paste0(save_name, "_comp_bar.png"),plot = com_bar,width = 20,height = 10,dpi = 500)
  
  print(com_bar)
}
```

Gene Set Enrichment Analysis with Cluster Profiler

```{r}
library(clusterProfiler)
library(enrichplot)
organism = "org.Hs.eg.db"
#BiocManager::install(organism,character.only = TRUE)
library(organism,character.only = TRUE)
```

```{r}
GSEA_plot = function(res,comp_name, ref_cond, test_cond){
  
  original_gene_list = res$log2FoldChange
  
  names(original_gene_list) = rownames(res)
  
  gene_list = na.omit(original_gene_list)
  
  gene_list = sort(gene_list,decreasing = TRUE)
  
  names(gene_list) = gsub("\\..*", "", names(gene_list))
  
  gene_list = gene_list[!duplicated(names(gene_list))]
  
  gse = gseGO(geneList = gene_list,
              ont = "BP",
              keyType = "ENSEMBL",
              minGSSize = 3,
              maxGSSize = 800,
              pvalueCutoff = 0.05,
              verbose = TRUE,
              OrgDb = org.Hs.eg.db,
              pAdjustMethod = "BH")
  
  require(DOSE)
  
  gse@result$.sign <- ifelse(
    gse@result$NES > 0,
    paste0("Activated in ", test_cond),
    paste0("Activated in ", ref_cond)
  )
  
  gse@result$.sign <- factor(gse@result$.sign)
  
  GSEA_dotplot = dotplot(gse,showCategory = 10, split=".sign") + facet_grid(.~.sign) +
    ggtitle(paste0("GSEA Dot Plot for ", comp_name)) +
  theme(
        plot.title = element_text(size = 20),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(size = 12,angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12)
        ) 
  
  save_name = gsub(" ","_",comp_name)
  setwd(analysis_path)
  ggsave(filename = paste0(save_name, "_GSEA_dotplot.png"),plot = GSEA_dotplot,width = 20,height = 10,dpi = 500)
  print(GSEA_dotplot)
}

```


