---
title: "RNA-Seq Analysis"
output: html_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Running the DESeq2 Analysis.

Load the necessary libraries for the analysis.
```{r}
library(tidyverse)
library(tximport)
library(DESeq2)
library(readr)
#library(tximportData)
```

Set the sample names and the conditions.
### CHANGE SAMPLE NAMES AND CONDITIONS HERE FOR DIFFERENT ANALYSES ###

```{r}
# Add in the sample names as a list below
sample_names <- c('input sample names here')
# Add in the conditions below with the correct number of replicates for each condition
sample_condition <- c(rep('Condition 1',3),rep('Condition 2',3),rep('Condition 3',3))
unique_condition = unique(sample_condition)
num_conditions = length(unique_condition)

```


Define the abundance files paths and import the tx2gene tsv. 
### CHANGE PATHS HERE FOR DIFFERENT ANALYSES ###

```{r}
analysis_path = "/path/to/kallito/outputs"
abundance_path = file.path(analysis_path, sample_names,"abundance.tsv")
names(abundance_path) = sample_names

tx2gene = read.table("/path/to/tx2gene/file/tx2gene.clean.tsv",sep = "\t")

```

Run tximport on the abundance files. Maps transcript level counts to gene level counts for DESeq2 analysis.

```{r}
txi_samples <- tximport(abundance_path, type="kallisto", tx2gene=tx2gene)
```

define the sample dataframe and make a deseq2 dataframe. 

```{r}
sample_df <- data.frame(sample = factor(sample_names),
              condition = factor(sample_condition,levels = unique_condition ))
# set row names attribute of sample_df
row.names(sample_df) <- sample_names

dds <- DESeqDataSetFromTximport(txi_samples,
colData = sample_df,
design = ~condition)
```


filter out low count genes 

```{r}
# pre-filter low count genes
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```


Relevel and make multiple dds objects if there are more than 2 conditions

```{r}
# If there are greater than 2 conditions, this will create a list of dds objects in which a different condition is the reference 
if (num_conditions > 2){
  dds_list = c()
  for (c in 1:num_conditions) {
    dds$condition = relevel(ddscondition,ref = unique_condition[c])
    dds_c = DESeq(dds)
    dds_list[[ unique_condition[c] ]] <- dds_c
  }
  
  res_list <- c()
  res_shrink_list <- c()
  
  for (ref in unique_condition) {  
      dds_ref <- dds_list[[ref]]
      
      # Get coefficient names from DESeq2
      coef_names <- resultsNames(dds_ref)
      
      # Loop through coefficients
      for (cn in coef_names) {
          # Skip the intercept
          if (cn == "Intercept") next
          
          message("Processing: ", cn)
          
          # Extract result
          res <- results(dds_ref, name = cn, alpha = 0.05)
          
          # Run lfcShrink
          res_shrink <- lfcShrink(dds_ref, coef = cn, type = "apeglm")
          
          # Save with meaningful names
          res_list[[ cn ]] <- res
          res_shrink_list[[ cn ]] <- res_shrink
      }
  }
  
  ensembl_names = row.names(res_shrunk_list[1])
  
} else {
  
  coef_name <- resultsNames(dds)[2]
  res <- results(dds, name = coef_name, alpha = 0.05)
  
  res_shrunk <- lfcShrink(dds,
                          coef = coef_name,
                          type = "apeglm")
  
  ensembl_names = row.names(res_shrunk)
  
}
```


Get Gene symbol names for each ENSG ID:

```{r}
library(biomaRt)
```

Rid of version numbers in ENSG IDs and use Biomart to get Gene names. 

```{r}
# rids of version numbers after a "." character in ENSG IDs
ensembl_names_no_version <- gsub("\\.\\d+$", "", ensembl_names)

# searches biomart for external gene names 
ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")

gene_info <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                       filters = "ensembl_gene_id",
                       values = ensembl_names_no_version,
                       mart = ensembl)

```

```{r}
# create a vector of gene names aligned with res
gene_names <- gene_info$external_gene_name[match(ensembl_names_no_version, gene_info$ensembl_gene_id)]

# replace NAs with the Ensembl ID itself
gene_names[is.na(gene_names)] <- ensembl_names[is.na(gene_names)]
```


Give every res object a column with external gene name

```{r}
if (num_conditions > 2) {
  
  for (res in res_shrink_list){
    res$geneID = gene_names
  }
  
} else {
  res_shrunk$geneID = gene_names
}

```

Function for volcano plot and results 

```{r}
volcano_plot = function(res,lfc_val,comp_name){
  
  res$log2FoldChange = as.numeric(res$log2FoldChange)
  # Keep NA as NA, do not replace
  res$neglog_padj <- -log10(res$padj)
  # Optionally, remove rows with NA if you want to completely exclude them from plotting
  res_plot <- res[!is.na(res$neglog_padj), ]
  res_plot$significant = ifelse(!is.na(res_plot$padj) & res_plot$padj < 0.05, "Significant", "Not Significant")  

  plot_labels = res_plot[!is.na(res_plot$padj) & abs(res_plot$log2FoldChange) > lfc_val & res_plot$padj < 0.05, ]
  
  volcano_plot = ggplot(res_plot, aes(x = log2FoldChange, y = neglog_padj, color = significant)) +
    geom_point(size = 0.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "blue") +
    geom_vline(xintercept = lfc_val, linetype = "dotted", color = "blue") +
    geom_vline(xintercept = lfc_val*(-1), linetype = "dotted", color = "blue") +
    geom_text(data = plot_labels, aes(label = geneID, vjust = -0.5), size = 2) +
    annotate("text", x = Inf, y = Inf, label = paste0(length(res_plot$geneID)," genes present"), hjust = 2, vjust = 1.1, size = 3) +
  xlim(-10,10)+
  labs(title = paste0("Volcano Plot for ", comp_name), x = "Log2 Fold change", y = "-log(adjusted p-value)", 
       color = "Significance of Differential Expression")
  
    # Saves the volcano plot to a directory
  setwd(analysis_path)
  ggsave(paste0(comp_name,"_volcanoplot.png"), plot = volcano_plot, width = 10, height = 6, dpi = 300)
  
  res_top_genes = res[!is.na(res$padj) & abs(res$log2FoldChange) > 1 & res$padj < 0.05, ]
  res_top_genes_sorted = res_top_genes[order(res_top_genes$padj,decreasing = FALSE),]
  res_top_10 = head(res_top_genes_sorted,10)
  res_top_10 = as.data.frame(res_top_10)
  top_file_name = paste0(comp_name,collapse = "")
  write_csv(res_top_10,file = paste0(top_file_name,"_top10_genes.csv"))
  
  
  return(volcano_plot)
  
}
```


## Heat maps for specific genes

```{r}
library(pheatmap)
```

Function for generating heatmap based on genes of interest

```{r}
clustermap <- function(dds, genes, gene_names, plot_name){
  
  # Set rownames of dds to gene names (important if needed for indexing)
  rownames(dds) <- gene_names
  
  # Filter genes to only those present in gene_names
  valid_genes <- intersect(genes, gene_names)

  if (length(valid_genes) == 0) {
    stop("None of the input genes are present in gene_names.")
  }
  
  # Get size-factor normalized counts
  normvalues <- counts(dds, normalized = TRUE)

  # Subset to valid genes only
  gene_vals <- normvalues[valid_genes, , drop = FALSE]

  # Apply log2(x + 1)
  gene_vals_log <- log2(gene_vals + 1)

  # Plot
  setwd(analysis_path)
  png(paste0(plot_name, "_clustermap.png"), width = 5000, height = 5000, res = 500)

  pheatmap::pheatmap(
    gene_vals_log,
    show_rownames = TRUE,
    show_colnames = TRUE,
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    display_numbers = FALSE,
    cellwidth = 50,
    cellheight = 10,
    fontsize = 12,
    border_color = NA,
    main = paste0(plot_name, " Clustered Log2 Expression")
  )

  dev.off()
}
```


```{r}
int_heatmap = function(dds,genes,gene_names,plot_name){
  
  rownames(dds) = gene_names
  
  normvalues = counts(dds,normalized = T)
  
  gene_vals = normvalues[genes,]
  
  setwd(analysis_path)
  png(paste0(plot_name,"_heatmap.png"),width = 3000, height = 2000, res = 300)

  # Generate clustered heatmap
  pheatmap(gene_vals,
         show_rownames = TRUE,     
         show_colnames = TRUE,    
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         display_numbers = TRUE,
         cellwidth = 50,
         cellheight = 50,
         fontsize = 12,
         main = paste0(plot_name, " Gene Expression"))

  dev.off()
  
  
}
```

Example of executing cluster map function for given genes. 

```{r}
iMG_names = c("IL4","IL10","TMEM119","P2RY12","GPR34","CX3CR1","C1QA","STAT1","STAT6","CD68","IBA1","KLF4","KLF5","KLF6","NOS2","CD14","CD34","TGFB1","TNFA","IL12A","IL23A","CD11B")
clustermap(dds3,iMG_names,gene_names,"iMG")
```


